---
title: 'ATAC-seq: differential accessability analysis'
author: "Sam Buckberry"
output: html_document
---
---
title: "Untitled"
output: html_document
---

```{r}
library(GenomicRanges)
library(GenomicAlignments)
library(edgeR)

library(stringr)
library(magrittr)
library(ggplot2)
library(ggrepel)
```

Read the peak counts
```{r}
peak_counts <- read.table("processed_data/atac_peak_counts.tsv")
```

Set the treatment and control groups
```{r}
groups <- ifelse(test = grepl(pattern = "_T.U", x = colnames(peak_counts)),
                 yes = "Unstimulated", no = "Stimulated")
samples <- str_sub(string = colnames(peak_counts), start = 2, end = 5)
samples <- str_c(samples, groups, sep = "_")

```

Load counts into a DGEList
```{r}
y <- DGEList(counts = peak_counts, group = groups, samples = samples)
```

Design matrix for differential testing
```{r}
design <- model.matrix(~ 0 + groups)
design

# Clean up the column names (will make things easier later)
colnames(design) <- c("Stimulated", "Unstimulated")
design
```


Filter to remove peaks with low counts that would not be useful in statistical testing
```{r}
keep <- filterByExpr(y)
table(keep)

y <- y[keep, , keep.lib.sizes=FALSE]
```

Calculate normalisation factors
```{r}
y <- calcNormFactors(y)
y$samples
```


Differential testing using Limma-Voom
```{r}
v <- voom(counts = y, design = design, plot = TRUE)
```

What is voom doing?

Counts are transformed to log2 counts per million reads (CPM), where “per million reads” is defined based on the normalization factors we calculated earlier.
A linear model is fitted to the log2 CPM for each peak, and the residuals are calculated.
A smoothed curve is fitted to the sqrt(residual standard deviation) by average peak counts (see red line in plot above)
The smoothed curve is used to obtain weights for each peak and sample that are passed into limma along with the log2 CPMs.
More details for the application of this method to RNA-seq data can be found here.  https://genomebiology.biomedcentral.com/articles/10.1186/gb-2014-15-2-r29


```{r, fig.height=5}
boxplot(v$E, names=v$targets$samples)
```
If the boxplots look markedly different between samples (the above look ok), a quantile normalisation might be beneficial.
```{r}
vq <- voom(counts = y, design = design, plot = TRUE,
           normalize.method = "quantile")
```



```{r, fig.height=5}
boxplot(vq$E, names=v$targets$samples)
```


### Differential expression testing

Fitting linear models in Limma
```{r}
fit <- lmFit(object = v, design = design)
```

Make the contrasts matrix
```{r}
contr <- makeContrasts(Stimulated - Unstimulated, levels = colnames(coef(fit)))
contr
```

Estimate contrasts for each gene
```{r}
fit2 <- contrasts.fit(fit, contr)
```

Empirical Bayes smoothing of standard errors (shrinks standard errors that are much larger or smaller than those from other peaks towards the average standard error) (see https://www.degruyter.com/doi/10.2202/1544-6115.1027)
```{r}
eb <- eBayes(fit2)
```

Check out the top differential expressed genes
```{r}
topTable(eb)
```

Put all DE testing results into an object for plotting and saving
```{r}
tt2 <- topTable(eb, number = nrow(v))
tt2$Significant <- ((abs(tt2$logFC) > 1) & (tt2$adj.P.Val < 0.05))
```

How many DE peaks are there?
```{r}
table(tt2$Significant)
```
